---
title: "inclass_example2"
author: "Elisa"
date: "1/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# specify 50 bins
```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(bins = 50)

```


# draw density curve instead (take all values and draw smooth curve to represent proportions)
```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_density()
```


# you can also manipulate bandwidth (bw) but no example shown


# jitter spreads out points by adding random noise so you can see them more clearly (which is ok here bc the x-axis is not meaningful beyond discrete continent categories). otherwise, it would just be one straight line
```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_jitter()
```

# boxplot
```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_boxplot()
```


# boxplot AND jitter on top of each other
```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_jitter()
```

# ridges
```{r}
ggplot(gapminder, aes(lifeExp, continent)) +
  ggridges::geom_density_ridges()
```


# bar plots
```{r}
ggplot(mtcars, aes(cyl)) +
  geom_bar()
```
# ggplot assumes cylinder is a continuous variable, so there are extra values on y-axis

# bar plots with geom_col
```{r}
ggplot(mtcars, aes(cyl, count)) +
  geom_col()
```
# for geom_col, you must specify x and y axes
# not sure why it's not working


# line charts
```{r}
tsibble::as_tsibble(co2) %>% 
  rename(yearmonth = index,
         conc = value) %>% 
  mutate(month = lubridate::month(yearmonth, label = TRUE),
         year  = lubridate::year(yearmonth)) %>% 
  ggplot(aes(month, conc)) +
  geom_line(aes(group = year), alpha = 0.5) +
  ylab("CO2 Concentration")
```





