---
title: "inclass_examples"
author: "Elisa"
date: "1/30/2020"
output: html_document
---

Scatterplot with ggplot

# use gapminder data, assign x = gdpPercap and y = lifeExp
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp))
```


# the graph is blank because we've only given it 2 out of 7 essential parts. we need to tell it which geometric object to use to plot points

# look at the list of possible aesthetics
?geom_point

# added geom_point to specify that we want a scatterplot
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point()
```

# we have overplotted, all the points blend together

# added alpha = 0.1 to specify what fraction of completely solid this is
# alpha = 0.1 specifies that you need 10 points to get to a solid black color
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1)
```

# add color outside of aes function, which will make all points blue
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1, color = "blue")
```

# the scale of x-axis isn't very informative, so we can add a log transformation

```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita")
```
# you don't need to name the arguments if you put them in order

# here is an example with the arguments explicitly named (added "name =" GDP per capita)
# if you named the first one, you have to name all of them. for example, if you named "data = gapminder", then you must name the following arguments too.
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10(name = "GDP per capita")
```

# next, tell it to use a dollar format for the x-axis variable (GDP per capita)
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format())
```

# scale here means you're pulling from the "scales" package
# :: means "look for a function in a specific package (in this case, scales)"
# sometimes different packages will use the same name; for instance, the "filter" function in dplyr and base r

# add a theme to make it less ugly AND change y label
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  theme_bw() +
  ylab("Life Expectancy")
```

# change breaks for scale of lifeExp
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  scale_y_continuous(breaks = c(10, 20, 30, 40, 50, 60)) +
  theme_bw() +
  ylab("Life Expectancy")
```

# make lifeExp scale begin at 0 (ylim)
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  scale_y_continuous(breaks = c(10, 20, 30, 40, 50, 60)) + ylim(0, 200) +
  theme_bw() +
  ylab("Life Expectancy")
```

# make lifeExp scale begin at 0 (ylim)
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) + ylim(0, 100) +
  theme_bw() +
  ylab("Life Expectancy")
```

# not sure if this worked...




# to add color by continent, add it INSIDE the aes function
```{r scatterplot with ggplot}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(mapping = aes(color = continent), alpha = 0.1)
```





