---
title: "File Input/Output"
author: "Elisa"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in libraries
```{r load libraries}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(tsibble))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(haven))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(svglite))
```

Load in gapminder dataset
```{r}
(gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia"))
```
Note: Putting the whole thing in parentheses will have the same result as the print() function.

```{r}
write_csv(gap_asia_2007, "exported_file.csv")
```

Check working directory
```{r}
getwd()
```

Relative file path, according to working directory
```{r}
file.path("data", "cats_data.csv")
```

Save dataset in folders within working directory
```{r}
write_csv(gap_asia_2007, here::here("data", "s008_data", "exported_file.csv"))
```

Imported csv file into R again by specifying the path where it exists.
```{r}
read_csv(here::here("data", "s008_data", "exported_file.csv"))
```

Read file from internet
```{r}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
```

Import excel file from online. First, create the necessary folders so you the specified file path in the next step exists.
```{r}
dir.create(here::here("data", "s008_data"), recursive = TRUE)
```
Recursive = TRUE means create any folders along the way. If recursive = FALSE, then it will only create the newest folder.

Import dataset from online.
```{r}
xls_url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/GreatestGivers.xls"
download.file(xls_url, here::here("data", "s008_data", "some_file.xls"), mode = "wb")
```
Note: "some_file.xls" is not recommended as a file name.

Load in clevel data using haven package
```{r}
(clevel <- haven::read_spss(here::here("data", "s008_data", "clevel.sav")))
```

Clean clevel data
Language and gender variables are currently SPSS format (numeric), but we want them as factors in R. Here, we convert them.
```{r}
clevel_cleaned <-
  clevel %>% 
  mutate(language = as_factor(language),
         gender = as_factor(gender),
         isClevel = factor(isClevel, 
                           levels = c(0, 1), 
                           labels = c("No", "Yes"))
  ) %>% 
  print()
```

Save cleaned file as a new CSV file so it's smaller and easier to import again in the future.
```{r}
write_csv(clevel_cleaned, here::here("data", "s008_data", "clevel_cleaned.csv"))
```

Make a plot
```{r}
clevel_plot <-
  clevel_cleaned %>% 
  mutate(isClevel = recode(isClevel, 
                           No = "Below C-level", 
                           Yes = "C-level"),
         gender = recode(gender,
                         Female = "Women",
                         Male = "Men")) %>% 
  ggplot(aes(paste(isClevel, gender, sep = "\n"), Extraversion, color = gender)) +
  geom_boxplot() +
  geom_jitter(height = .2) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  ggtitle("Extraversion Stan Scores") +
  scale_y_continuous(breaks = 1:9) +
  ggthemes::theme_fivethirtyeight()

print(clevel_plot)
```


Use ggsave function to save plot. .svg is the preferable format, but it's not working due to some error, so we're saving as a .jpeg here.
```{r}
dir.create(here::here("output", "figures"), recursive = TRUE)
ggsave(here::here("output", "figures", "clevel_extraversion.jpeg"), clevel_plot)
```


