---
title: "File Input/Output"
author: "Elisa"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in libraries
```{r load libraries}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(tsibble))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(readxl))
```

Load in gapminder dataset
```{r}
(gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia"))
```
Note: Putting the whole thing in parentheses will have the same result as the print() function.

```{r}
write_csv(gap_asia_2007, "exported_file.csv")
```

Check working directory
```{r}
getwd()
```

Relative file path, according to working directory
```{r}
file.path("data", "cats_data.csv")
```

Save dataset in folders within working directory
```{r}
write_csv(gap_asia_2007, here::here("data", "s008_data", "exported_file.csv"))
```

Imported csv file into R again by specifying the path where it exists.
```{r}
read_csv(here::here("data", "s008_data", "exported_file.csv"))
```

Read file from internet
```{r}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
```

Import excel file from online. First, create the necessary folders so you the specified file path in the next step exists.
```{r}
dir.create(here::here("data", "s008_data"), recursive = TRUE)
```
Recursive = TRUE means create any folders along the way. If recursive = FALSE, then it will only create the newest folder.

Import dataset from online.
```{r}
xls_url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/GreatestGivers.xls"
download.file(xls_url, here::here("data", "s008_data", "some_file.xls"), mode = "wb")
```

Note: "some_file.xls" is not recommended as a file name.

