---
title: 's04: `dplyr` Exercise'
output: 
  html_document:
    keep_md: true
    theme: paper
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

**When you make an Rmd file for participation or homework, be sure to do this**:

1. Change the file output to both html and md _documents_ (not notebook).
  - See the `keep_md: TRUE` argument above.

2. `knit` the document. 

3. Stage and commit the Rmd and knitted documents.


# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages.
    - This loads `dplyr`, too.

Hint: You can add the `suppressPackageStartupMessages()` function around the 
      `library()` command to keep your output looking nice!
    
```{r load packages, warning = FALSE, message = FALSE}
# load your packages here:
library(gapminder)
library(tidyverse)
```
    

## `select()`

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from 
the `gapminder` data, in that order.

```{r}
select(gapminder, year, lifeExp, country)
# will be displayed in order that you name them in

# mtcars example
select(mtcars, cyl, disp)

```

2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)

# Better way:
select(gapminder, country:lifeExp)

# you can also do the following, but problems include ambiguity of which variables you intend to include (if someone else is reading it) and also, this is dependent on the order of your rows, so if you swap your rows, you might accidentally use a different variable
select(gapminder, 1:4)

# mtcars example
select(mtcars, mpg:drat)

```

3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, -lifeExp)

# mtcars example
select(mtcars, -hp)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
select(gapminder, continent, everything())

# mtcars example
select(mtcars, hp, everything())
```

5. Rename `continent` to `cont`.

```{r}
# compare
select(gapminder, cont = continent)

# rename and keep all other variables
select(gapminder, cont = continent, everything())

# keeps renamed variable in same place, doesn't change anything else
rename(gapminder, cont = continent)

# rename all variables to uppercase
rename_all(gapminder, toupper)
rename_all(gapminder, tolower)

# to save these changes, store them as a new object
gapminder_new <- rename(gapminder, cont = continent)
gapminder_new


# mtcars example
select(mtcars, MiPerGal = mpg)
select(mtcars, MiPerGal = mpg, everything())
rename(mtcars, MiPerGal = mpg)
```


## `arrange()`

1. Order by year.

```{r}
arrange(gapminder, year)
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, desc(year))

# can also use minus sign
arrange(gapminder, -year)
```

3. Order by year, then by life expectancy.

```{r}
arrange(gapminder, year, lifeExp)
```


## Piping, `%>%`

Note: think of `%>%` as the word "then"!

Demonstration:

Here I want to combine `select()` Task 1 with `arrange()` Task 3.

### Base R method

This is how I could do it by *nesting* the two function calls:

```{r nesting functions example, eval = FALSE}
# Nesting function calls can be hard to read
arrange(select(gapminder, year, lifeExp, country), year, lifeExp)
```

```{r intermediate assignment example, eval = FALSE}
gap_sel <- select(gapminder, year, lifeExp, country)
arrange(gap_sel, year, lifeExp)
```

### tidyverse method

Now using with pipes:

```{r}
# alter the below to include 2 "pipes"
arrange(select(gapminder, year, lifeExp, country), year, lifeExp)


gapminder %>%
  select(year, lifeExp, country) %>% 
  arrange(year, lifeExp)

```




# Back to Guide 

Return to guide at the section on *Relational/Comparison and Logical Operators in R*.


# Transforming datasets

## `filter()`

1. Only take data with population greater than 100 million.

```{r}
gapminder %>%
  filter(pop > 100000000)

# if you want to run analyses on this, do the following
gapminder %>%
  filter(pop > 100000000) %>% 
  analyses()

# can also create a new dataset
data_clean <- gapminder %>% 
  filter(pop > 200000) %>% 
  analyses()

```

2. Your turn: of those rows filtered from step 1., only take data from Asia.

```{r}
gapminder %>%
  filter(pop > 100000000) %>% 
  filter(continent == "Asia")

# this also works
gapminder %>%
  filter(pop > 100000000, continent == "Asia")

# this also works
gapminder %>%
  filter(pop > 100000000 & continent == "Asia")
```

3. Repeat 2, but take data from countries Brazil, and China. 

```{r}
gapminder %>%
  filter(pop > 100000000, country == "Brazil" | country == "China")

# can also use the %in% function, which will check whether each value for country is in our list of options, which only includes Brazil and China
gapminder %>%
  filter(pop > 100000000, country %in% c("Brazil", "China"))

# THIS WILL NOT YIELD THE SAME RESULT because it will compare one element at a time to each vector, then recycle. so is the first value equal to Brazil? next, is the second value equal to China? next, is the third value equal to Brazil? etc...
gapminder %>%
  filter(pop > 100000000, country == c("Brazil", "China"))

# Notes about this

# country == c("Brazil", "China") means the following:
# A = A (True or false)
# B = B (True or false)
# C = A (True or false)
# D = B (True or false)
# 
# country %in% c("Brazil", "China") means the following:
# A = A or B (True or false)
# B = A or B (True or false)
# C = A or B (True or false)
# D = A or B (True or false)

```


## `mutate()` (10 min)

The `mutate()` function _creates_ new columns in the tibble by transforming other variables. Like `select()`, `filter()`, and `arrange()`, the `mutate()` function also takes a tibble as its first argument, and returns a tibble. This function keeps all other variables.

The general syntax is:

```
mutate(tibble, NEW_COLUMN_NAME = CALCULATION)
```

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
# add step 1
gapminder %>%
  mutate(grossGDP = gdpPercap * pop)

# add step 2
gapminder %>%
  mutate(grossGDP = gdpPercap * pop,
         gdpBill = grossGDP/1000000000)

# add step 3
gapminder %>%
  mutate(grossGDP = gdpPercap * pop,
         gdpBill = round(grossGDP/1000000000, 2))
# can also specify "digits = 2"

# another way, using round() function
gapminder %>%
  mutate(grossGDP = gdpPercap * pop,
         gdpBill = (grossGDP/1000000000) %>% round(2))

# drop some variables (gdpPercap, pop, grossGDP)
gapminder %>%
  mutate(grossGDP = gdpPercap * pop,
         gdpBill = (grossGDP/1000000000) %>% round(2)) %>% 
           select(-gdpPercap, -pop, -grossGDP)
```

Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>%
  transmute(gdpBill = (pop * gdpPercap / 1000000000) %>%  round(2))

# if you want to keep some other variables, just call them first
gapminder %>%
  transmute(country, continent, lifeExp,
            gdpBill = (pop * gdpPercap / 1000000000) %>%  round(2))
```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
gapminder %>%
  mutate(lifeExp = if_else(country == "Canada" & year == 1952,
                           70,
                           lifeExp)) %>% 
  filter(country == "Canada")

# if_else function: if_else(condition, what happens if true, what happens if false)
# if_else(if country is Canada and year is 1952, then make the lifeExp value be 70. if this isn't true, stick with current lifeExp value)
# last line is just to view only Canada to check if it worked
```


Replace life expectancy for Bangladesh with NA_real_ but leave all other variables.

```{r}
gapminder %>%
  mutate(lifeExp = if_else(country == "Bangladesh",
                           NA_real_,
                           lifeExp)) %>% 
  filter(country == "Bangladesh")
```



Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
```

These functions we've seen are called __vectorized functions__—they are designed 
to work with vectors and do their computation on each element of the vector(s).

## git stuff

Now is a good time to knit, commit, push!


# Back to Guide Again

Let's head back to the guide at the section on `summarize()`.


# Exercises for grouped data frames

Let's do some practice with grouping (and ungrouping) and summarizing data frames!

1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(FILL_THIS_IN) %>% 
  FILL_THIS_IN(min_life = min(lifeExp))
```


2. Let's compute the mean Agreeableness score across items for each participant 
in the `psych::bfi` dataset. Be sure to handle `NA`!

```{r}
psych::bfi %>%
  as_tibble() %>% 
  select(A1:A5) %>% 
  rowwise() %>% 
  mutate(A_mean = mean(c(A1, A2, A3, A4, A5), na.rm = TRUE)) %>% 
  ungroup()
```

Now compute mean scores for the other Big Five traits, as well as 
`sd` and `min` scores for reach person.

```{r}

```

** There are a few other ways to do this sort of computation.**

`rowMeans()` computes the mean of each row of a data frame. We can use it by
putting `select()` inside of `mutate()`:

```{r}
psych::bfi %>% 
  as_tibble() %>% 
  select(A1:A5) %>% 
  mutate(A_mn = rowMeans(select(., A1:A5)),
         A_mn2 = rowMeans(select(., starts_with("A", ignore.case = FALSE))))
```

Some functions are **vectorized**, so you don't need `rowwise()`. 
For example, `pmin()` computes the "parallel min" across the vectors it receives:

```{r}
psych::bfi %>% 
  as_tibble() %>% 
  select(A1:A5) %>% 
  mutate(A_min = pmin(A1, A2, A3, A4, A5))
```


3. Calculate the growth in population since the first year on record 
_for each country_ by rearranging the following lines, and filling in the 
`FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
knitr::kable()
group_by(country) %>% 
```

```{r}

```


4. Determine the country, on each continent, that experienced the 
**sharpest 5-year drop in life expectancy**, sorted by the drop, by rearranging 
the following lines of code. Ensure there are no `NA`'s. A helpful function to 
compute changes in a variable across rows of data (e.g., for time-series data) 
is `tsibble::difference()`:

```
drop_na() %>% 
ungroup() %>% 
arrange(year) %>% 
filter(inc_life_exp == min(inc_life_exp)) %>% 
gapminder %>% 
mutate(inc_life_exp = FILL_THIS_IN) %>% # Compute the changes in life expectancy
arrange(inc_life_exp) %>% 
group_by(country) %>% 
group_by(continent) %>% 
knitr::kable()
```

```{r}

```


# Bonus Exercises

If there's time remaining, we'll practice with these three exercises. 
I'll give you a minute for each, then we'll go over the answer.

1. In `gapminder`, take all countries in Europe that have a GDP per capita 
   greater than 10000, and select all variables except `gdpPercap`. 
   (Hint: use `-`).

2. Take the first three columns of `gapminder` and extract the names.

3. In `gapminder`, convert the population to a number in billions.

4. Take the `iris` data frame and extract all columns that start with 
   the word "Petal". 
    - Hint: take a look at the "Select helpers" documentation by running the 
      following code: `?tidyselect::select_helpers`.

5. Filter the rows of `iris` for Sepal.Length >= 4.6 and Petal.Width >= 0.5.

Exercises 4. and 5. are from 
[r-exercises](https://www.r-exercises.com/2017/10/19/dplyr-basic-functions-exercises/).